---
title: "Building and validating a Soil Quality Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building and validating a Soil Quality Index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(soilquality)
```

A soil quality index is a pipeline of five decisions, and this package now
offers a choice at every one of them:

```
select  →  group  →  score  →  weight  →  aggregate  →  VALIDATE
```

The last step is the one most often skipped, and the one this vignette treats
as the point of the exercise. An index has **no ground truth**. Nothing tells
you whether yours is right. What you *can* establish is whether it
discriminates, and whether your conclusion survives building it differently.

We use `soil_structured`, which is simulated but carries the covariance real
soil has: texture summing to 100, organic carbon tied to organic matter,
exchange capacity generated by clay and organic colloids.

```{r}
props <- c("Sand", "Silt", "Clay", "BD", "pH", "OM", "SOC",
           "N", "P", "K", "CEC", "Ca", "Mg", "S", "EC")
str(soil_structured[, props[1:6]])
```

## Step 0. Is the data worth factoring at all?

Almost no published index checks this. It costs one line.

```{r}
pca_adequacy(soil_structured[, c("pH", "OM", "N", "P", "K", "CEC", "BD")])
```

Note what happens when the particle-size fractions are included: they sum to
100 by construction, the correlation matrix becomes singular, and KMO is
undefined. That is a property of compositional data, not a bug, and the
function says so rather than returning a number.

```{r}
pca_adequacy(soil_structured[, c("Sand", "Silt", "Clay", "pH", "OM")])$kmo_message
```

## Step 1. Select a minimum data set

Two routes, rewarding different things. PCA favours indicators carrying
**variance**; correlation-network analysis favours **hubs**.

```{r}
std <- standardize_numeric(soil_structured[, props])

pca_route <- pca_select_mds(std)
pca_route$mds
```

```{r eval = requireNamespace("igraph", quietly = TRUE)}
net_route <- na_select_mds(soil_structured[, props])
net_route$mds
```

They disagree completely here, and that disagreement is informative rather
than embarrassing: `mds_consensus()` returns the intersection, and an empty
intersection is a signal to look at the data before trusting either.

### The relative loading rule

`pca_select_mds()` takes one indicator per component by default. The rule
stated in the literature keeps everything within 10 % of the top loading,
which usually gives a larger set:

```{r}
length(pca_select_mds(std)$mds)
length(pca_select_mds(std, within = 0.10)$mds)
```

## Step 2. Group by function

Selecting across the whole pool lets one dominant function crowd out the
others. Here, ungrouped selection leaves entire functions unrepresented:

```{r}
groups <- assign_function_groups(props)
names(groups)

grouped <- pca_select_mds(std, groups = groups, selector = "norm")
grouped$mds
```

Group by **function**, not by physical/chemical/biological — Maaz et al.
(2023) tested that familiar split by confirmatory factor analysis and found it
has no statistical support.

## Step 3. Score

Linear is the package default. The sigmoidal curve dominates the literature:

```{r}
mds <- grouped$mds

linear_rules  <- standard_scoring_rules(mds, scoring = "linear")
sigmoid_rules <- standard_scoring_rules(mds, scoring = "sigmoid")

sigmoid_rules[[1]]
```

The literature contradicts itself on which is better, so compute both and
check whether your conclusion moves. That is Step 6.

### Scoring against a reference soil

Both of the above normalise against your own sample, which forces the best
site to score about 1 whatever its condition. Scoring against an undisturbed
reference is the documented escape — at the cost of needing a defensible
reference soil.

```{r}
reference_rules <- list(
  OM = reference_scoring(reference = 5.5),
  BD = reference_scoring(reference = 1.05, direction = "lower"),
  pH = reference_scoring(reference = 6.5, direction = "optimum",
                         tolerance = 1.5)
)
reference_rules$BD
```

## Steps 4 and 5. Weight and aggregate

Weighting is the most contested step in the whole pipeline, which is a good
reason to try skipping it. The area route ignores weights entirely.

```{r}
weighted <- compute_sqi_properties(
  soil_structured, properties = mds, id_column = "SampleID",
  scoring_rules = linear_rules, method = "weighted"
)

area <- compute_sqi_properties(
  soil_structured, properties = mds, id_column = "SampleID",
  scoring_rules = linear_rules, method = "area"
)

summary(weighted$results$SQI)
```

## Step 6. Validate — the point of all this

```{r}
validation <- sqi_validate(weighted)
validation
```

The **distribution across decision categories** is the headline, not the
correlation with anything. Maaz et al. found two indices correlating at
r = 0.96 while one put 94 % of plots in the middle band and the other 61 %.
An index that calls everything "medium" cannot inform a decision, however well
it correlates.

```{r fig.width = 7, fig.height = 4.5}
plot_sqi_validation(validation)
```

### Does the conclusion survive a change of recipe?

```{r}
sigmoid <- compute_sqi_properties(
  soil_structured, properties = mds, id_column = "SampleID",
  scoring_rules = sigmoid_rules
)

sqi_stability(linear = weighted, sigmoid = sigmoid, area = area)
```

A high rank correlation does not rescue a changed extreme. If the top field is
what you act on and it moves when you switch scoring functions, you do not
have a result — you have an artefact of the recipe.

### Fidelity to everything you measured

```{r}
tds <- compute_sqi_df(
  soil_structured[, c("SampleID", props)],
  id_column = "SampleID", select = "none"
)

sqi_validate(weighted, tds = tds)$fidelity
```

Read this carefully: the total-data-set index is **not ground truth**. It is
the index with everything in it. High fidelity means your reduced set is
faithful to your full measurement set, not that either is correct.

## A decision table

| Step | Options | Default | When to change it |
|---|---|---|---|
| Select | `pca_select_mds`, `na_select_mds`, expert list | PCA | Network when hubs matter more than variance, or normality is doubtful |
| Group | `groups = NULL`, `soil_function_groups` | none | Whenever more than one soil function matters |
| Score | linear, sigmoid, optimum, threshold, reference | linear | Sigmoid to reproduce most published work; reference for cross-study comparability |
| Weight | AHP, loading, centrality, equal | equal | AHP when you have defensible expert judgement |
| Aggregate | `"weighted"`, `"area"` | weighted | Area to sidestep weighting entirely |
| Validate | `sqi_validate`, `sqi_stability` | — | Always |

## One warning worth repeating

Do not build an index from **predicted** properties. Chaudhry et al. (2024)
found that computing an SQI from spectrally predicted properties gave
R² = 0.23, while predicting the index directly from the same spectra gave
R² = 0.90 — with individually acceptable property models. If you only have
predictions, model the index itself.
